---
title: "F1 Race Wins Analysis"
author: "Fasih Atif (2001385)"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

F1 is a sport that is loved by millions of fans each year who love to watch it for its speed, technology, rivalries, and drama. Formula 1 recently completed its 70 years anniversary and last weekend ended the 2020 season with the Grand Finale at Abu Dhabi. Lewis Hamilton clinched his 7th world title equaling Michael Schumachers record. This got me thinking over who have been the best F1 drivers over the past 70 years. The metric that i chose to to measure success was number of race wins.

```{r, echo = FALSE, message= FALSE,warning = FALSE}
local({r <- getOption("repos")
       r["CRAN"] <- "http://cran.r-project.org" 
       options(repos=r)
})

library(tidyverse)
library(data.table)
library(rvest)
library(gganimate)
library(stringr)
install.packages("gifski")
library(gifski)
library(knitr)
```

```{r, echo = FALSE, message= FALSE,warning = FALSE}

# Web scraping data via rvest
# Data cleaning done inside the function with if else statement

race_results <- function(x){
  
url <- read_html(x)


#### Pull race names
gp_name <- url %>%
  html_nodes('.limiter+ .bold') %>%
  html_text()
if (length(gp_name) ==0) {
  gp_name <- ''}
  else {
    # fix race names
    gp_name <- trimws(gsub("\n","",gp_name))
  }

  
#### Pull race dates
race_date <- url %>%
  html_nodes('.dark.hide-for-mobile') %>%
  html_text()
if (length(race_date) == 0) {
  race_date <- ''
}
 

#### Pull race winners
driver_winner <- url %>%
  html_nodes('.dark+.bold') %>%
  html_text()
if (length(driver_winner) ==0) {
  driver_winner <- ''}
else{
  # Fix race winner names
  driver_winner <- trimws(gsub("\n","",driver_winner)) # remove \n from string
  driver_winner <- gsub("\\s+"," ",driver_winner) # remove all empty spaces between words
  driver_winner <- gsub('.{4}$', '', driver_winner) # remove last 4 characters
}


#### Pull team names
team <- url %>%
  html_nodes('.uppercase') %>%
  html_text()
if (length(team) ==0) {
  team  <- ''}
else{
  # Fix team name
  team<- team[seq(2, length(team), 2)] # extract every 2nd element from vector
}
  

#### Pull no of laps
laps <- url %>%
  html_nodes('.bold.hide-for-mobile') %>%
  html_text()
if (length(laps) ==0) {
  laps  <- ''
}

# Join data in dataframe
df <- data.frame('race_name' = gp_name, 'race_date' = race_date, 'winner' = driver_winner,'team' = team, 'laps' = laps)
return(df)
write_csv(f1_results.csv)
}
```

```{r, echo = FALSE, message= FALSE,warning = FALSE}
#### Apply lapply function
get_all_seasons <- function(my_url){
  database <- rbindlist(lapply(my_url,race_results))
  database <<- database
  return(database)
}
```

```{r,echo = FALSE, message= FALSE,warning = FALSE}
#### Set for loop to get URLs
get_all_seasons_results <- function(from_year,to_year){
i <- 1
my_urls <- list()
for(page_result in seq(from = from_year, to = to_year, by = 1)) {
  my_urls[i] = paste0('https://www.formula1.com/en/results.html/',page_result,'/races.html')
  i <- i+1
}
get_all_seasons(my_urls) #call function to input urls in lapply
}
```

```{r, echo = FALSE, message= FALSE,warning = FALSE}
# Function with date parmeters to pull results
# get_all_seasons_results(from_year,to_year)
get_all_seasons_results(2015,2020)
```

Our dataframe for the race results looks like below:

```{r, echo = FALSE, message= FALSE,warning = FALSE}
table_format <- tail(database)
library(knitr)
kable(table_format)


```

## Exploratory Data Analysis

### Top 5 most no of wins by F1 drivers (1950-2020)

We can check who have been the most successful drivers of all time using the visualization below:

```{r, fig.width = 8, fig.height = 5 , echo = FALSE, message= FALSE,warning = FALSE}

database %>% 
  group_by(winner) %>%
  summarize(total_wins = n()) %>%
  arrange(desc(total_wins)) %>%
  head(5) %>%
  ggplot(aes(x = reorder(winner, total_wins), y = total_wins, fill = winner,label = total_wins)) + geom_col() + coord_flip() +geom_text(hjust = -0.3) +
  labs( x = "F1 Race Winners", y = "Total no of Wins", title = "Top 5 most no of wins by F1 drivers (2010-2020)")
  
```

The graph shows that Lewis Hamilton has been the most successful F1 driver of all time with 95 wins followed by the renowned Michael Schumacher with 91 wins and Sebastian Vettel with 53 wins.

```{r, echo = FALSE, message= FALSE,warning = FALSE}
database$year <-str_sub(database$race_date, start= -4)


```

```{r, ECHO = FALSE, message= FALSE,warning = FALSE}
database %>% 
  group_by(race_name) %>%
  summarize(most_venues = n()) %>%
  arrange(desc(most_venues)) %>%
  head(5) %>%
  ggplot(aes(x = reorder(race_name, most_venues), y = most_venues, fill = race_name,label = most_venues)) + geom_col() + coord_flip() +geom_text(hjust = -0.3) +
  labs( x = "F1 Race Winners", y = "Total no of Wins", title = "Top 5 most no of wins by F1 drivers (2010-2020)")
```

## Top 5 most successful drivers winning progression

We previously calculated the top 5 most successful drivers of all time. Now we will focus at the rate at which these 5 drivers gathered their race wins over the their racing careers:

```{r, echo = FALSE, fig.width = 11, fig.height=5.5, message= FALSE,warning = FALSE}

#Filter database for top 5 most successful drivers
drivers <- c("Lewis Hamilton","Michael Schumacher", "Sebastian Vettel", "Alain Prost", "Ayrton Senna")
top_5_stats<- database %>% filter(winner %in% drivers)

# Convert year column to int
top_5_stats <- top_5_stats %>% mutate(year = as.numeric(year))

# Aggregate no of wins for each year for the 5 drivers
top_5_per_year<- top_5_stats %>% 
  group_by(year,winner) %>%
  count(winner)

# Calculate the cumulative sum of the wins per year for each driver
top_5_per_year_cs <- top_5_per_year %>% group_by(winner) %>% arrange(winner,year) %>% mutate(cs = cumsum(n))

# Visualize the no of wins for each driver over their racing career  
top_5_per_year_cs %>% ggplot(aes(x = year, y = cs, fill = winner,group = winner)) + geom_line(aes(colour=winner),size = 2) +
  labs( x = "Year", y = "Races Won", title = "Top 5 most successful drivers winning progression") + 
  scale_y_continuous(breaks = seq(0,100, by=20)) + 
  scale_x_continuous(breaks = seq(1978,2020, by=4)) + 
  theme(legend.position = "none",axis.title=element_text(size=14,face="bold")) +     theme(axis.text = element_text(size = 14), plot.title = element_text(size = 20, face = "bold",vjust = 2.5,hjust = 0.5)) +
    annotate("text", x = 1994, y = 57, label = "Alain Prost", fontface = 2) +
    annotate("text", x = 1995, y = 45, label = "Ayrton Senna", fontface = 2) +
    annotate("text", x = 2008, y = 96, label = "Michael Schumacher", fontface = 2) +
    annotate("text", x = 2018, y = 99, label = "Lewis Hamilton", fontface = 2) +
    annotate("text", x = 2019, y = 40, label = "Sebastian Vettel", fontface = 2)
  
  
```

Lewis Hamilton has carried an impressive form from 2014 onward with a steep increase in race wins. Michael Schumachers race wins rate is nearly identical to Lewis Hamilton but with an occasional slow down in some periods. Both had careers ranging approximately 12 years over which they amassed over 90 wins each. Sebestian Vettel had a better start compared to his rival Lewis Hamilton bagging a lot wins and consequently championships from 2010-2013. From 2014 onwards with his move to Ferrari, the rate of race wins decreased significantly. Alain Prost gathered 51 race wins over his 11 year racing period at an average of 5 race wins a season. Ayrton Senna had a slow start with 6 wins only in his first 4 years but then displayed dominant performance over the next 6 years until his death during a race crash.

**Racing Bar Chart Visualization for Most no of Wins by F1 Drivers (2015-2020)**

We will now take a subset of our results for 2015-2020 from our results database and visualize it using a racing bar chart to show how drivers rankings for most wins have changed over past 5 years.

```{r, echo = FALSE, message= FALSE,warning = FALSE}

# Import a table of cumulative race wins for the past 5 years
# This table was made using our original results database

#driversTable <- read_csv('cumulative_results.csv',col_names = TRUE)

my_url_cs <- "https://raw.githubusercontent.com/fasihatif/web_scraping/master/final_project/cumulative_results.csv"
driversTable <- read_csv(my_url_cs)

# Prepare data for ranking
win_ranking <- driversTable %>%
  group_by(year) %>%
  # The * 1 makes it possible to have non-integer ranks while sliding
  mutate(rank = min_rank(-value) * 1,
         Value_rel = value/value[rank==1],
         Value_lbl = paste0(" ",value)) %>%
  filter(rank <=10) %>%
  arrange(value,winner) %>%
  ungroup() %>%
  group_by(year) %>%
  arrange(rank, winner) %>%
  mutate(rank = seq(1, n())) %>%
  ungroup() 

# Build individual plots for each year
plots_year <- ggplot(win_ranking, aes(rank, group = winner, 
                     fill = as.factor(winner), color = as.factor(winner))) +
  geom_tile(aes(y = value/2,
                height = value,
                width = 0.9), alpha = 0.8, color = NA) +
  geom_text(aes(y = 0, label = paste(winner, " ")), vjust = 0.2, hjust = 1) +
  geom_text(aes(y=value,label = Value_lbl, hjust=0),size = 7) +
  coord_flip(clip = "off", expand = FALSE) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_reverse() +
  guides(color = FALSE, fill = FALSE) +
  
  labs(title = "Total no of wins by F1 Drivers Since {closest_state}", x = "", y = "Winners",
       caption = "Most No of wins by F1 Drivers 2015-2020") +
  theme(plot.title = element_text(hjust = 0, size = 22),
        axis.ticks.y = element_blank(),  # These relate to the axes post-flip
        axis.text.y  = element_blank(),  # These relate to the axes post-flip
        plot.margin = margin(1,1,1,4, "cm")) +
  
  transition_states(year, transition_length = 4, state_length = 1) +
  ease_aes('cubic-in-out')
      

# Save as preferred rendered format
# It will take a minute or so to create GIF
gif <- animate(plots_year, 200, fps = 5, duration = 13, width = 800, height = 600, renderer = gifski_renderer("gganim.gif"))

```

```{r,echo = FALSE, message= FALSE,warning = FALSE}
knitr::include_graphics(gif)
```
